### 2023-02-02 manuscript.Rmd
add Mayo BB data

Proposed journal: ***Molecular Neurodegeneration***

@@

Associations between the Chromosome 19 CAA risk locus, CpG methylation, and CAA pathology are consistent with the effect of this locus on disease being mediated by DNA methylation. Furthermore, the same two CpG sites are associated with dementia in an independent cohort, indicating that genomic signals other than *APOE* $\epsilon$ alleles in the *APOE* region may independently affect dementia risk. The atherosclerosis risk variant rs2000660 is in strong linkage disequilibrium with a synonymous coding variant (rs650724) of *COL4A1*, providing a candidate functional variant. *BIN1* is associated with neurofibrillary tangles and neuritic plaques but not with diffuse amyloid pathology. *TMEM106B* is associated with hippocampal sclerosis and TDP-43 pathology but not the canonical Alzheimer's disease pathologies. Our findings provide novel insights into neuropathology risk factors and parse clinical Alzheimer's genetic risk into specific key neuropathological domains.

@@

The most common form of AD, late-onset AD (LOAD), presents in individuals aged 65 years and older and is highly heritable, with twin and family studies estimating its heritability at \~60% [@ridge2013; @gatz1997; @gatz2006]. Genome-wide association studies (GWAS) have to date identified over 70 genetic risk loci for LOAD that are involved biological processes including amyloid precursor protein processing, immune response, and extracellular matrix regulation [@jansen2019; @wightman2021; @bellenguez2022]. Most dementia GWAS to date have focused on clinically diagnosed LOAD or proxy outcomes based on family history of dementia [@jansen2019; @wightman2021; @bellenguez2022; @kunkle2019]; however, late-onset dementia diagnosed as LOAD is increasingly recognized as a heterogeneous clinical syndrome that can reflect multiple underlying independent or co-occurring pathological processes that may have independent or pleiotropic genetic risk factors [@farfel2016; @katsumata2022; @karanth2020].

@@

In addition to proteinopathies, non-stroke vascular pathologies are common in elderly autopsied research participants and contribute to cognitive decline and dementia [@skrobot2016]. Cerebral amyloid angiopathy (CAA) is one such vascular pathology characterized by amyloid-beta deposition in cerebral blood vessels [@weber2018]. CAA often co-occurs with LOAD, but can independently contribute to cerebral injury by causing hemorrhages in the brain parenchyma [@weber2018; @skrobot2016]. Infarcts of both small but grossly visible lacunar arteries and microscopically examined vessels (the latter referred to as microinfarcts) are also common in aged individuals and contribute to cognitive decline [@smith2012; @skrobot2016]. Cerebral atherosclerosis and sclerosis of small cerebral blood vessels, called brain arteriolosclerosis, predispose individuals to infarcts and hippocampal sclerosis [@arvanitakis2017; @neltner2014]; moreover, they contribute to cognitive decline even after adjusting for the presence of related pathologies [@arvanitakis2016; @ighodaro2017]. Collectively, these factors reveal an increasingly complex and synergistic web of pathologies contributing to cognitive impairment and dementia, and demonstrate the need for continued study into genomic risk factors of dementia beyond those of clinically diagnosed LOAD.

The investigation of genetic risk of neuropathologies is an attractive approach for uncovering biological pathways involved in the development of LOAD and related dementias. Neuropathologies are commonly operationalized in aged autopsy cohorts as binary or semi-quantitative (or ordinal) neuropathology endophenotypes (NPE). Prior GWAS of neuropathology endophenotypes have confirmed known LOAD risk loci and have identified potential neuropathology risk loci [@beecham2014; @shade2022; @vattathil2021; @farrell2022]. However, continued recruitment in ongoing autopsy cohorts and methodological development for GWAS and downstream functional analysis provides additional opportunity for in-depth investigation into multiple neuropathologies studied across multiple cohorts.

@@

We used genotype and neuropathology data from four high-quality data sources: the National Alzheimer's Coordinating Center (NACC), the Religious Orders Study and the Rush Memory and Aging Project (ROSMAP), the Adult Changes in Thought (ACT) study, and the AD Neuroimaging Initiative (ADNI). Some participants in ROSMAP and ADNI also had neuropathology and genotype data available in NACC. In these cases, records in the NACC were preferentially kept in order to maximize sample size of the initial analysis using only NACC participants (see Statistical Analyses subsection).

@@

@@

@@

## Single-variant analysis identifies novel neuropathology-associated mapped to *COL4A1* and *PIK3R5*

@@

In total, 14 independent ($r^2 < 0.05$) loci had lead variants that met the genome-wide significance threshold of $p < 5 \times 10^{-8}$ in the main pooled mega-analysis. Of these, four were duplicate signals in the *APOE* region, leaving 10 significant signals across 11 endophenotypes. Novel variant-endophenotype associations included (1) a variant 12 kbp upstream of *COL4A1,* rs2000660, associated with atherosclerosis in the circle of Willis (**Figure 1**) and (2) a *PIK3R5* intronic variant, rs72807981, associated with Braak NFT stage (**Figure 2**). Additionally, one locus with lead variant rs4803778 remained significantly associated with CAA in the *APOE* $\epsilon$-adjusted analysis (**Figure 3**). Loci identified that have been previously associated with NPE or dementia included *APOE* with Braak stage, CAA, CERAD score, diffuse amyloid plaques, and LATE-NC; *BIN1* with Braak NFT stage; and *TMEM106B* with HS and LATE-NC (**Table 2**).

@@

Visual inspection of a dendrogram (**Supplementary Figure S1**) of neuropathology endophenotypes based on polychoric correlations reveals two primary clusters of NPE: (1) a cluster of AD- or amyloid-associated endophenotypes consisting of Braak NFT stage, CERAD score, diffuse amyloid, and CAA; and (2) a cluster of vascular endophenotypes consisting of atherosclerosis, arteriolosclerosis, gross infarcts, and microinfarcts. PC1 in the AD cluster accounted for 66% of variance in the cluster, while PC1 in the vascular cluster accounted for 50% of the variance in that cluster. Only the *APOE* region was significantly associated with the AD cluster PC1, while no loci were significantly associated with the vascular PC1.

## Colocalization analyses identifies colocalization between CAA and DNA methylation in *APOC2* region

@@

Multiple suggestive NPE loci showed evidence of colocalization with eQTL in GTEx. A locus within *GRN* suggestively associated with HS ($p = 6.9 \times 10^{-8}$) colocalized with *GRN* expression ($\text{PrC}>99.9\%$), and the *TMEM106B* locus associated with HS and LATE-NC colocalized with *TMEM106B* eQTL. The CAA-associated *APOC2* locus independent of *APOE* $\epsilon$ diplotype also colocalized with *APOE* and *APOC2* eQTL in GTEx ($\text{PrC}= 95\%$) and with four mQTL in ROSMAP ($\text{PrC}= 97\%$ for each of the four mQTL, see **Figure 3**).

@@

We performed GWAS and downstream in silico functional analyses for eleven neuropathology endophenotypes across four high-quality neuropathology cohorts with a maximum sample size of 7,463 participants. Our work builds on previous genetic association studies of NPE, Alzheimer's, and related dementias and provides another attempt to better understand the complex associations between different neuropathologies and genetic risk factors [@katsumata2022; @shade2022; @beecham2014; @dugan2021; @bellenguez2022; @vattathil2021]. Our study has several important limitations, including sample size and heterogeneity in study design. However, we discovered intriguing new loci mapped to *COL4A1* and *PIK3R5* associated with atherosclerosis in the circle of Willis and Braak stage for neurofibrillary tangles, respectively. We also investigated known loci in *BIN1*, *APOE*, and *TMEM106B* to provide additional context on their associations with multiple NPE.

## Limitations and conclusion

There are limitations to the approach we took in this study. The four data sources we used, while all of high quality, used different study designs and recruitment strategies. NACC participant data arises from recruitment at each ADRC and likely consists of a more clinical population, whereas MAP and ACT participants may be more representative of the broader aged communities from which they are recruited. ROS participants include Catholic brothers, sisters, and priests and may represent a population more highly educated than the United States as a whole. These differences are reflected in the relative frequencies and distributions of NPE between cohorts, with NACC and ADNI tending to show more severe pathology than ACT and ROSMAP, and their mean ages of death were markedly lower (**Tables 1 and S2**). Additionally, there is likely substantial heterogeneity in the way that neuropathology data were collected and graded both within and between studies. We review these potential limitations briefly in our previous GWAS of brain arteriolosclerosis [@shade2022].

Additionally, while recruitment, autopsy, and genotyping of the cohorts used is ongoing, the sample sizes available for genome-wide investigation of NPE genetic risk is small relative to their clinical counterparts. A recent GWAS of AD included over 111,000 clinically diagnosed or "proxy" (based on family history) AD cases and nearly 700,000 controls. In contrast, our largest analyzed sample size for CERAD score totaled 7,447 participants, while LATE-NC totaled only 2,224. These sample reflect immense progress in autopsy cohorts from even eight years ago and will only continue to increase, but available sample sizes still pose limitations on power of GWAS and similar 'omics approaches to studying disease risk using neuropathology endophenotypes. A strength of our study is that data on multiple neuropathology is available for each participant, allowing for multivariate approaches to studying NPE. However, as our PCA-based multivariate analyses did not identify any risk loci not identified in single-outcome analyses, this approach alone is not sufficient overcome sample size limitations.

@@

One locus on Chromosome 12q34 with lead variant rs2000660 (minor allele frequency $[\text{MAF}]= 9.9\%$ in pooled data set) located 12 kbp upstream of *COL4A1* was significantly associated with atherosclerosis in the circle of Willis (odds ratio $[\text{OR}] =0.71$, $p = 3.9 \times 10^{-8}$). *COL4A1* in the brain is preferentially expressed in endothelial cells (**Figure 1)** and codes for a component of collagen IV, an important component of basal lamina. In previous GWAS, researchers have reported the *COL4A1/COL4A2* locus to be associated with numerous other vascular phenotypes, including peripheral artery disease, coronary artery disease, stroke, and arteriolar stiffness [@klarin2019, @nikpay2015, @steffensen2018]. The *COL4A1/COL4A2* locus has also been found to be associated with rare familial cerebrovascular diseases and lacunar ischemic stroke[@blevins2021; @rannikmae2017]. In a recent GWAS, rs2000660 in particular was a risk variant for migraines [@hautakangas2022]. The relevance of the *COL4A1* locus to cerebral vascular traits is thus highly supported by previous research, and the biological role of collagen IV to vascular disease is possibly related to disruption of the extracellular matrix [@steffensen2018].

rs2000660 was not a QTL in GTEx or ROSMAP. However, rs650724, a variant in high LD with rs2000660 ($r^2=0.84$), is a synonymous coding variant (p.Ser1600Ser in ENST00000375820.10; p.Ser319Ser in ENST00000650424.1) for *COL4A1*. rs2000660 was not nominally associated with any other vascular NPE in our study, and a previous GWAS of cerebral atherosclerosis using ROSMAP participants did not identify the *COL4A1* as a risk locus [@vattathil2021]. However, the sample size of the previous study was significantly smaller than the one used in the present study (1,325 vs 6,959). Indeed, in the present study, rs2000660 reached genome-wide significance only in the pooled mega-analysis, though it was nominally significant in the ROSMAP-only validation analysis ($p=0.0079$).

@@

The other identified novel genome-wide significant locus located on Chromosome 17p13 was associated with Braak NFT stage and had a lead variant of rs72807981 ($\text{OR} =0.71$, $p = 3.9 \times 10^{-8}$, $\text{MAF}= 6.3\%$), an intronic variant within *PIK3R5*, which codes for a phosphatidylinositol 3-kinase involved in cell growth, motility, and survival. This variant was suggestively associated with Braak NFT stage in NACC ($\text{OR} =0.69$, $p = 1.8 \times 10^{-7}$) and was nominally validated in ACT ($\text{OR} =0.50$, $p = 0.028$). The same variant was also nominally associated with neuritic plaques in the pooled analysis ($\text{OR} =0.82$, $p = 0.0047$) but was not nominally associated with any other AD NPE. rs72807981 is not a QTL in GTEx or ROSMAP, nor were any of the variants in high LD with it in our study. There is previous research suggesting that *PIK3R5* may be involved in the development of NFT. One study compared differential mRNA expression in aged adults with Braak NFT stage VI vs. non-demented controls and found *PIK3R5* to be more highly expressed in cases than controls (false-discovery rate $[FDR]= 0.01$) [@guennewig2021]. Interestingly, *PIK3R5* is expressed preferentially in microglial cells in humans (see **Figure 2)**, suggesting that its association with neurofibrillary pathology may be immune-mediated [@zhang2016].

## *APOE* and NPE

Multiple variants in the *APOE* region were associated with pathognomonic AD NPE, including Braak stage, CERAD score, diffuse amyloid plaques, and CAA. Variants in the *APOE* region were also associated with LATE-NC, which is consistent with previous genetic association studies of NPE [@beecham2014; @dugan2021]. While multiple loci in this region were significantly associated with each of these four NPE, only one CAA-associated locus with lead variant rs4803778 remained significant ($\text{OR} = 1.24$, $p = 5.8 \times 10^{-12}$) when adjusting for *APOE* $\epsilon$ diplotype. rs4803778 was in high LD ($r^2 = 0.97$) with rs4803779, which was also significantly associated with CAA ($\text{OR} = 1.24$, $p = 6.7 \times 10^{-12}$), is a lead mQTL in ROSMAP for five methylation sites ($p = 3.3 \times 10^{-91}$ for top site cg09555818; see **Figure 3**), and colocalizes with four of these mQTL ($\text{PrC} = 97\%$ for all four sites). Furthermore, two of these associated methylation sites, including cg09555818, were significantly associated with dementia in the Generation Scotland: Scottish Family Health Study [@walker2020]. These results indicate that this locus may be associated with CAA through epigenetic regulation. However, this locus is located approximately 30 kbp downstream of the *APOE* coding end site and is situated between the *APOC4*, *APOC2*, and *CLPTM1*. While *APOE* may be the most likely target gene of this locus due to its known association with CAA, this inference is not straightforward. In GTEx, this locus is a stronger eQTL of *APOC2* than *APOE*, although it is strongly significant and colocalizes with expression of both genes in multiple tissues. Future work should examine the regulatory role of the identified methylation sites and test them for association with CAA and other AD NPE.

## HS and LATE-NC

One intronic locus of *TMEM106B* was significantly associated with both HS ($\text{OR} = 1.50$, $p = 1.5 \times 10^{-11}$) and LATE-NC ($\text{OR}= 1.50$, $p = 1.5 \times 10^{-11}$). Additionally, a locus within *GRN* was suggestively associated with HS ($\text{OR} = 0.72$, $p = 6.9 \times 10^{-8}$). Both of these genes have been found previously to be associated with frontotemporal lobar degeneration [@rollinson2011; @ciani2019], and in a recent GWAS by Bellenguez et al., both of these genes were identified as risk factors for LOAD [@bellenguez2022a]. We downloaded available summary statistics from the GWAS Catalog and performed additional colocalization analyses at these loci to investigate if these loci were shared between clinical LOAD phenotypes and NPE [@buniello2019]. We found that HS ($\text{PrC}= 91\%$) and LATE-NC ($\text{PrC}= 91\%$) colocalized with ADRD at both the *TMEM106B* (HS and LATE-NC both $\text{PrC}= 91\%$) and *GRN* (HS $\text{PrC}\gt 99.9\%$, LATE-NC $\text{PrC}= 90\%$) loci (**Figure 4**). HS and LATE-NC also colocalized with each other at these loci, as discussed above. These results indicate that HS, LATE-NC, and LOAD likely share causal loci at these genes.

## *BIN1*, neurofibrillary tangles, and neuritic plaques

>>>>

Proposed journal: **???**

@@

Associations between the Chromosome 19 CAA risk locus, CpG methylation, and CAA pathology are consistent with the effect of this locus on disease being mediated by DNA methylation. Furthermore, the same two CpG sites are associated with dementia risk factors in independent cohorts, indicating that genomic signals other than *APOE* $\epsilon$ alleles in the *APOE* region may independently affect dementia risk. The atherosclerosis risk variant rs2000660 is in strong linkage disequilibrium with a synonymous coding variant (rs650724) of *COL4A1*, providing a candidate functional variant. *BIN1* is associated with neurofibrillary tangles and neuritic plaques but not with diffuse amyloid pathology. *TMEM106B* is associated with hippocampal sclerosis and TDP-43 pathology but not the canonical Alzheimer's disease pathologies. Our findings provide novel insights into neuropathology risk factors and parse clinical Alzheimer's genetic risk into specific key neuropathological domains.

@@

The most common form of AD, late-onset AD (LOAD), presents in individuals aged 65 years and older and is highly heritable, with twin and family studies estimating its heritability at \~60% [@ridge2013; @gatz1997; @gatz2006]. Genome-wide association studies (GWAS) have to date identified over 70 genetic risk loci for LOAD that are involved biological processes including amyloid precursor protein processing, tau, and immune response [@jansen2019; @wightman2021; @bellenguez2022]. Most dementia GWAS to date have focused on clinically diagnosed LOAD or proxy outcomes based on family history of dementia [@jansen2019; @wightman2021; @bellenguez2022; @kunkle2019]; however, late-onset dementia diagnosed as LOAD is increasingly recognized as a heterogeneous clinical syndrome that can reflect multiple underlying independent or co-occurring pathological processes that may have independent or pleiotropic genetic risk factors [@farfel2016; @katsumata2022; @karanth2020].

@@

In addition to proteinopathies, non-stroke vascular pathologies are common in elderly autopsied research participants and contribute to cognitive decline and dementia [@skrobot2016]. Cerebral amyloid angiopathy (CAA) is one such vascular pathology characterized by amyloid-beta deposition in cerebral blood vessels [@weber2018]. CAA often co-occurs with LOAD, but can independently contribute to cerebral injury by causing hemorrhages in the brain parenchyma [@weber2018; @skrobot2016]. Infarcts of both small but grossly visible lacunar arteries and microscopically examined vessels (the latter referred to as microinfarcts) are also common in aged individuals and contribute to cognitive decline [@smith2012; @skrobot2016]. Cerebral atherosclerosis and sclerosis of small cerebral blood vessels, called brain arteriolosclerosis, predispose individuals to infarcts and hippocampal sclerosis [@arvanitakis2017; @neltner2014]; moreover, they contribute to cognitive decline even after adjusting for the presence of related pathologies [@arvanitakis2016; @ighodaro2017]. Collectively, these factors reveal an increasingly complex and synergistic web of pathologies which contribute to cognitive impairment and dementia.

```{=html}
<!-- Insert paragraph on how genetic risk factors of clinical alzheimer's may be associated with specific neuropathologies and pathways, and how studying the genomic risk factors of individual neuropathologies may provide a finer grained understanding of the genetic risk of AD

and they demonstrate the need for continued study into genomic risk factors of dementia beyond those of clinically diagnosed LOAD.-->
```

The reality of clinically diagnosed AD being a heterogeneous catch-all clinical syndrome with multiple and varied pathologies found in the brains of those diagnosed has implications for the interpretation of AD GWAS (which can often be more accurately described as dementia GWAS). Each individual risk loci identified in large AD GWAS may be associated with some brain pathologies but not others. If this is so, we would expect that investigation into the pathologies themselves may provide a reasonable alternative approach to discovering relevant risk genes and biological pathways involved in the development of LOAD and related dementias. Prior GWAS of neuropathology endophenotypes (NPE) have confirmed known LOAD risk loci and have identified potential neuropathology risk loci [@beecham2014; @shade2022; @vattathil2021; @farrell2022]. However, continued recruitment in ongoing autopsy cohorts and methodological development for GWAS and downstream functional analysis provides additional opportunity for in-depth investigation into multiple neuropathologies studied across multiple cohorts.

@@

We used genotype and neuropathology data from four high-quality data sources: the National Alzheimer's Coordinating Center (NACC), the Religious Orders Study and the Rush Memory and Aging Project (ROSMAP), the Adult Changes in Thought (ACT) study, and the AD Neuroimaging Initiative (ADNI). Some participants in ROSMAP and ADNI also had neuropathology and genotype data available in NACC. In these cases, records in the NACC were preferentially kept in order to maximize sample size of the initial analysis using only NACC participants (see Statistical Analyses subsection). An overview of our study design is presented in **Supplementary Figure S1**.

@@

## DNA Methylation Data

Pre-processed and quality-controlled DNA methylation data for 740 ROSMAP participants were downloaded from Synapse.org (Synapse IDs: syn3157275 and syn3191087). DNA methylation preparation and quality control measures have been previously described [@yu2015; @dejager2018]. Briefly, approximately 50mg of frozen gray matter tissue from the dorsolateral prefrontal cortex were sampled from each participant. DNA was then extracted and processed using the Illumina Infinium HumanMethylation450 BeadChip. Quality control measured included remove low-quality probes; removing participants with poor bisulfite-conversion efficiency; and adjusting methylation levels by age, sex, and batch, which adequately controlled for batch effects. Missing methylation levels were imputed using 100-nearest neighbors [@yu2015; @dejager2018].

@@

## Association between CpG site methylation and cerebral amyloid angiopathy

In our *APOE* $\epsilon$-adjusted genetic association analysis, one locus near *APOE* remained significantly associated with cerebral amyloid angiopathy. We found that this locus colocalized with DNA methylation levels at four CpG sites in ROSMAP. To investigate whether these CpG sites were themselves association with CAA pathology, we combined connected individual-level DNA methylation data and neuropathological data in ROSMAP for analysis. We used cumulative logit models using the "clm" function implemented in the "ordinal" R package with the semi-quantitative CAA variable described above as the outcome for analysis. We performed four analyses, with one of each of the four CpG sites tested as the independent variable of interest in each analsyis. We adjusted for age, sex, ROS vs MAP study, bisulfite conversion efficiency, post-mortem interval, and *APOE* $\epsilon$ genotype in each analysis. Z-values from the resulting parameter estimates were used as test statistics for statistical significance.

@@

## GWAS identifies novel neuropathology-associated variants mapped to *APOC2,* *COL4A1,* and *PIK3R5*

@@

In total, 14 independent ($r^2 < 0.05$) loci had lead variants that met the genome-wide significance threshold of $p < 5 \times 10^{-8}$ in the main pooled mega-analysis. Of these, four were duplicate signals in the *APOE* region, leaving 10 significant signals across 11 endophenotypes. Novel variant-endophenotype associations included (1) a variant 12 kbp upstream of *COL4A1,* rs2000660, associated with atherosclerosis in the circle of Willis (**Figure 1**) and (2) a *PIK3R5* intronic variant, rs72807981, associated with Braak NFT stage (**Figure 2**). Additionally, one locus between *APOC2* and *CLPTM1* with lead variant rs4803778 remained significantly associated with CAA in the *APOE* $\epsilon$-adjusted analysis (**Figure 3a**). Loci identified that have been previously associated with NPE or dementia included *APOE* with Braak stage, CAA, CERAD score, diffuse amyloid plaques, and LATE-NC; *BIN1* with Braak NFT stage; and *TMEM106B* with HS and LATE-NC (**Table 2**).

@@

Visual inspection of a dendrogram (**Supplementary Figure S2**) of neuropathology endophenotypes based on polychoric correlations reveals two primary clusters of NPE: (1) a cluster of AD- or amyloid-associated endophenotypes consisting of Braak NFT stage, CERAD score, diffuse amyloid, and CAA; and (2) a cluster of vascular endophenotypes consisting of atherosclerosis, arteriolosclerosis, gross infarcts, and microinfarcts. PC1 in the AD cluster accounted for 66% of variance in the cluster, while PC1 in the vascular cluster accounted for 50% of the variance in that cluster. Only the *APOE* region was significantly associated with the AD cluster PC1, while no loci were significantly associated with the vascular PC1.

## Colocalization analyses identifies colocalization between cerebral amyloid angiopathy, *APOC2* expression, and CpG methylation near *APOC2* in prefrontal cortex

@@

Multiple suggestive NPE loci showed evidence of colocalization with eQTL in GTEx. A locus within *GRN* suggestively associated with HS ($p = 6.9 \times 10^{-8}$) colocalized with *GRN* expression ($\text{PrC}>99.9\%$), and the *TMEM106B* locus associated with HS and LATE-NC colocalized with *TMEM106B* eQTL. In total, 42 NPE loci colocalized with QTL with $\text{PrC}= \ge80\%$, with 271 total colocalizing NPE-QTL pairs across 43 tissues (**Supplementary Table S4**).

The *APOE* $\epsilon$-independent Chromosome 19 CAA-associated locus colocalized with expression of multiple genes in GTEx: *APOE* expression in the wall of the aorta; *8CLPTM1* expression in the skin of the leg and suprapubic region; and *APOC2* expression in 14 different tissues, including the brain cortex, caudate, nuclues accumbens, and cerebellum ($\text{PrC}= 90-97\%$). Additionally, the same locus colocalized with mQTL for four CpG sites in ROSMAP, cg09555818, cg04401876, cg10169327, and cg13119609 ($\text{PrC}= 97\%$, see **Figure 3b**).

## DNA methylation levels at CpG sites cg09555818 and cg13119609 in the cerebral cortex are associated with cerebral amyloid angiopathy

In total, 708 ROSMAP participants had DNA methylation and CAA data available for analysis. We test methylation levels at each of the four CpG sites that colocalize with the *APOE* $\epsilon$-independent CAA risk locus on Chromosome 19 -- cg09555818, cg04401876, cg10169327, and cg13119609 -- for association with CAA pathology. Of these, hypomethylation of cg09555818 (odds ratio $[\text{OR}] =0.82$, $p = 0.004$) and cg13119609 ($\text{OR} =0.78$, $p = 0.0007$) were significantly associated with worse CAA pathology (see **Figure 3c**).

@@

We performed GWAS and downstream in silico functional analyses for eleven neuropathology endophenotypes across four high-quality neuropathology cohorts with a maximum sample size of 7,463 participants. Our work builds on previous genetic association studies of NPE, Alzheimer's, and related dementias and provides another attempt to better understand the complex associations between different neuropathologies and genetic risk factors [@katsumata2022; @shade2022; @beecham2014; @dugan2021; @bellenguez2022; @vattathil2021]. Our study has several important limitations, including sample size and heterogeneity in study design. However, we identified a novel APOE $\epsilon$-independent CAA risk locus which also affects *APOC2* brain expression and dementia-associated CpG methylation sites, which in turn were also associated with CAA risk. We also discovered intriguing new loci mapped to *COL4A1* and *PIK3R5* associated with atherosclerosis in the circle of Willis and Braak stage for neurofibrillary tangles, respectively. Lastly, we investigated known loci in *BIN1*, *APOE*, and *TMEM106B* to provide additional context on their associations with multiple NPE.

## Limitations

There are limitations to the approach we took in this study. The four data sources we used, while all of high quality, used different study designs and recruitment strategies. NACC participant data arises from recruitment at each ADRC and likely consists of a more clinical population, whereas MAP and ACT participants may be more representative of the broader aged communities from which they are recruited. ROS participants include Catholic brothers, sisters, and priests and may represent a population more highly educated than the United States as a whole. These differences are reflected in the relative frequencies and distributions of NPE between cohorts, with NACC and ADNI tending to show more severe pathology than ACT and ROSMAP, and their mean ages of death were markedly lower (**Tables 1 and S2**). Additionally, there is likely substantial heterogeneity in the way that neuropathology data were collected and graded both within and between studies. We review these potential limitations briefly in our previous GWAS of brain arteriolosclerosis [@shade2022]. Additionally, while recruitment, autopsy, and genotyping of the cohorts used is ongoing, the sample sizes available for genome-wide investigation of NPE genetic risk is modest relative to their clinical and proxy outcomes.

## Known AD risk loci are associated with specific neuropathological features

Multiple variants in the *APOE* region were associated with pathognomonic AD NPE, including Braak stage, CERAD score, diffuse amyloid plaques, and CAA. Variants in the *APOE* region were also associated with LATE-NC, which is consistent with previous genetic association studies of NPE [@beecham2014; @dugan2021]. Notably, *APOE* was not associated with vascular pathology except for CAA.

A locus approximately 30 kbp downstream of *BIN1* on Chromosome 2q14 was significantly associated with Braak stage ($\text{OR} = 1.21$, $p = 3.8 \times 10^{-10}$) and suggestively associated with CERAD score for neuritic plaques ($\text{OR} = 1.20$, $p = 6.6 \times 10^{-8}$). We verified through colocalization analysis that the same locus drives association signals with Braak NFT stage and CERAD score. In prior GWAS, this locus is second only to *APOE* for strength of association with LOAD [@bellenguez2022]. Interestingly, the lead variant in this locus, rs6733839, was associated with neither diffuse amyloid plaques ($\text{OR} = 1.05$, $p = 0.15$) nor CAA ($\text{OR} = 1.02$, $p = 0.49$). Previous research supports the hypothesis that *BIN1* is associated with LOAD through its effect on NFT rather than amyloid pathology [@holler2014; @franzmeier2019]. As neuritic plaques contain dying nerve cell processes with aberrant tau fibrils identical to those seen in neurofibrillary tangles [@nelson2009], our findings are also consistent with the hypothesis that *BIN1* influences AD risk primarily through tau rather than amyloid pathogenic processes.

One intronic locus of *TMEM106B* was significantly associated with both HS ($\text{OR} = 1.50$, $p = 1.5 \times 10^{-11}$) and LATE-NC ($\text{OR}= 1.50$, $p = 1.5 \times 10^{-11}$). Additionally, a locus within *GRN* was suggestively associated with HS ($\text{OR} = 0.72$, $p = 6.9 \times 10^{-8}$). Both of these genes are associated with frontotemporal lobar degeneration [@rollinson2011; @ciani2019], and with LOAD [@bellenguez2022a]. We downloaded available summary statistics from the GWAS Catalog and performed additional colocalization analyses at these loci to investigate if these loci were shared between clinical LOAD phenotypes and NPE [@buniello2019]. We found that HS, LATE-NC, and LOAD all colocalize at these two loci ($\text{PrC}\ge 90\%$, **Figure 4**). These results indicate that HS, LATE-NC, and LOAD likely share causal loci for these genes.

# *APOE* $\epsilon$-independent *APOC2* is associated with CAA

When we re-ran Chromosome 19 analyses on these phenotypes while adjusting for *APOE* $\epsilon$ genotype, only one CAA-associated locus with lead variant rs4803778 remained significant ($\text{OR} = 1.24$, $p = 5.8 \times 10^{-12}$). Several variants in this locus were lead eQTL for *APOC2* brain expression in GTEx and mQTL in ROSMAP for four methylation sites ($p = 3.3 \times 10^{-91}$ for top site cg09555818; see **Figure 3**). Colocalization analysis confirmed that the CAA risk locus shares a functional variants with both *APOC2* eQTL ($\text{PrC} = 89-97\%$ in brain tissues) and nearby mQTL ($\text{PrC} = 97\%$ for all four CpG sites). We confirmed that two of the CpG sites affected by the CAA risk locus, cg09555818 ($\text{OR} =0.82$, $p = 0.004$) and cg13119609 ($\text{OR} =0.78$, $p = 0.0007$), were in turn significantly associated with CAA pathology. Both of these CpG sites are located within the *APOC4-APOC2* readthrough transcript region overlapping *APOC4* and *APOC2*. We then confirmed that the directions of effects for rs4803779 risk allele, CpG methylation, and CAA risk were consistent with the hypothesis that CpG methylation mediates the genotype-phenotype association. Our results are consistent with the hypothesis that rs4803779 locus affects CAA risk through hypomethylation of CpG sites in the *APOC4-APOC2* region. These results are also consistent with previous studies in other human cohorts that implicate hypomethylation at cg09555818 and cg13119609 as potentially associated with Alzheimer's disease [@walker2021; @walker2020; @shao2018]. Circumstantial evidence indicates that *APOC2* may be the target gene of the rs4803779 risk locus as it colocalizes with *APOC2* expression in multiple brain tissues in GTEx and the associated CpG sites are located in exon 3 of *APOC4-APOC2*. However, more research will need to be done to verify that *APOC2* is the target gene.

@@

One locus on Chromosome 12q34 with lead variant rs2000660 (minor allele frequency $[\text{MAF}]= 9.9\%$ in pooled data set) located 12 kbp upstream of *COL4A1* was significantly associated with atherosclerosis in the circle of Willis ( $\text{OR} =0.71$, $p = 3.9 \times 10^{-8}$). *COL4A1* in the brain is preferentially expressed in endothelial cells (**Figure 1)** and codes for a component of collagen IV, an important component of basal lamina. rs2000660 was not nominally associated with any other vascular NPE in our study, and a previous GWAS of cerebral atherosclerosis using ROSMAP participants did not identify the *COL4A1* as a risk locus [@vattathil2021]. However, the sample size of the previous study was significantly smaller than the one used in the present study (1,325 vs 6,959). Indeed, in the present study, rs2000660 reached genome-wide significance only in the pooled mega-analysis, though it was nominally significant in the ROSMAP-only validation analysis ($p=0.0079$).

rs2000660 was not a QTL in GTEx or ROSMAP. However, rs650724, a variant in high LD with rs2000660 ($r^2=0.84$), is a synonymous coding variant (p.Ser1600Ser in ENST00000375820.10; p.Ser319Ser in ENST00000650424.1) for *COL4A1*. As rs2000660 does not have an obvious functional role and synonymous variants in some genes have been shown to alter mRNA stability, protein conformation, and other regulatory functions [@sauna2011], rs650724 presents as a possible functional variant driving genotype-atheroscerosis assosiation in this locus. In previous GWAS, researchers have reported the *COL4A1/COL4A2* locus to be associated with numerous other vascular phenotypes, including peripheral artery disease, coronary artery disease, stroke, and arteriolar stiffness [@klarin2019, @nikpay2015, @steffensen2018]. The *COL4A1/COL4A2* locus has also been found to be associated with rare familial cerebrovascular diseases and lacunar ischemic stroke[@blevins2021; @rannikmae2017]. In a recent GWAS, rs2000660 in particular was a risk variant for migraines [@hautakangas2022]. The relevance of the *COL4A1* locus to cerebral vascular traits is thus highly supported by previous research, and the biological role of collagen IV to vascular disease is possibly related to disruption of the extracellular matrix [@steffensen2018].

@@

The other identified novel genome-wide significant locus located on Chromosome 17p13 was associated with Braak NFT stage and had a lead variant of rs72807981 ($\text{OR} =0.71$, $p = 3.9 \times 10^{-8}$, $\text{MAF}= 6.3\%$), an intronic variant within *PIK3R5*, which codes for a phosphatidylinositol 3-kinase involved in cell growth, motility, and survival. This variant was suggestively associated with Braak NFT stage in NACC ($\text{OR} =0.69$, $p = 1.8 \times 10^{-7}$) and was nominally validated in ACT ($\text{OR} =0.50$, $p = 0.028$). The same variant was also nominally associated with neuritic plaques in the pooled analysis ($\text{OR} =0.82$, $p = 0.0047$) but was not nominally associated with any other AD NPE. rs72807981 is not a QTL in GTEx or ROSMAP, nor were any of the variants in high LD with it in our study. There is previous research suggesting that *PIK3R5* may be involved in the development of NFT. One study compared differential mRNA expression in aged adults with Braak NFT stage VI vs. non-demented controls and found *PIK3R5* to be more highly expressed in cases than controls (false-discovery rate $[FDR]= 0.01$) [@guennewig2021]. Interestingly, *PIK3R5* is expressed preferentially in microglial cells in humans (see **Figure 2**), suggesting that its association with neurofibrillary pathology may be immune-mediated [@zhang2016].
